---
title: "rQTL for ovule mutant mapping in *Mimulus nudatus*"
author: "Elen Oneal"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

```{r setup, include=FALSE,eval=TRUE}
require("knitr")
library(rmarkdown)
library(here)
```

```{r wrap-hook,include=FALSE,eval=TRUE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```

## Rationale

An ovule mutant phenotype was discovered in an inbred line of nudatus (GMR2), characterized by an early arrest of the ovule integument and the later development of the ovule primordia into filamentous structures. 

An F2 mapping population was created using the ovule mutant and DHRO22, a wildtype inbred line of M. nudatus. Both inbred lines were inbred for at least 6 generations. The F1 was selfed to generate F2s. Phenotyping of F2s revealed that the ovule mutant appears to segregate at a 25-30% rate.

F2.18 was encoded as both mutant and wildtype but didn't have enough sequence data to be included in the linkage map. It is not included in the QTL mapping.

Chunks marked eval=FALSE require input files generated by the shell pipeline documented in Ovules.Rmd. Pre-computed RDS files are provided in data/ so the plotting and model comparison chunks can be run independently.

## Prepare rQTL input files

The filtered phased genotype matrix provided by LepMAP3 and phenotype calls are reformatted into the CSV format required by rQTL. Genotype codes (AA, BB, AB) are converted to rQTL notation (A, B, H). Missing data is encoded as `-`. Output files are written to `data/`.

## Model choice
The binary model in rQTL is used in scanone and scantwo analyses because the ovule phenotype is scored as a binary trait (mutant vs wildtype). The Haley-Knott regression method is used for an initial fast scan to identify candidate QTL peaks, followed by the more computationally intensive EM algorithm for the final scan and permutation tests.

In the compare_models chunk below, we fit a glm model (y ~ Q1 + Q2 + Q1\*Q2) to determine whether a two-QTL interaction is present. If significant, this would indicate that there are at least two loci contributing to the ovule mutant phenotype and that they interact epistatically. Neither Q2 nor Q1*Q2 are significant in the fitted model, suggesting that the data do not support a two-QTL model.

```{r,create_rqtl,include=FALSE,eval=FALSE}
# Load required library for data manipulation.
suppressMessages(library(dplyr))

# Read the filtered genotype matrix produced earlier in the workflow.
ovule.geno <- read.table(here::here('data','ovule_sub_genotypes_missing_lowhet.txt'),h=T,sep='\t')

# Ensure position is numeric so sorting and downstream operations work correctly.
ovule.geno$position <- as.numeric(ovule.geno$position)

# Order markers by linkage group and genetic map position.
ovule.qtl <- ovule.geno[order(ovule.geno$lg,ovule.geno$cM),]

# Create a simplified marker ID, remove redundant location columns, and keep the genotype matrix.
ovule.qtl <- ovule.qtl %>%
  mutate(newmarker = paste('D', lg, sep='')) %>%
  mutate(newmarker = paste(newmarker, 'M', sep='')) %>%
  mutate(newmarker = paste(newmarker, position, sep='')) %>%
  mutate(marker = newmarker) %>%
  select(-newmarker, -position, -contig) %>%
  select(marker, lg, cM, everything())
ovule.qtl <- ovule.qtl[order(ovule.qtl$lg, ovule.qtl$cM),]
ovule.qtl$lg <- as.numeric(ovule.qtl$lg)

# Read phenotype calls and map sample IDs to F2 names.
phenotypes <- read.table(here::here('data','FinalPhenotypeCalls.txt'),h=T,sep='\t')
sample.names <- read.table(here::here('data','ovule_f2_samples.txt'),h=F)
sample.names <- sample.names %>% select(-V1)
colnames(sample.names) <- c('PlantID','F2')
phenotypes <- inner_join(sample.names, phenotypes, by='PlantID')
phenotypes <- phenotypes %>% select(-PlantID)
colnames(phenotypes)[2] <- 'phenotype'

# Define the genotype encoding used in the input data.
homo1 <- 'AA'
homo2 <- 'BB'
het <- 'AB'

# Function to convert the genotype matrix into QTL input files.
create_qtl_datasheet <- function(y, n, pheno, homo1, homo2, het){
  # Transpose the genotype matrix so samples become columns.
  y <- t(y)

  # Extract the sample names, dropping metadata rows.
  x <- rownames(y)
  x <- x[-(1:3)]
  x <- c('id', '', '', x)

  # Reattach the sample IDs as the first column.
  y <- cbind(x, y)
  rownames(y) <- c()
  colnames(y)[1] <- c('id')

  # Convert genotype codes to A/B/H/N notation used by rQTL.
  y <- gsub(homo1, 'A', y)
  y <- gsub(homo2, 'B', y)
  y <- gsub(het, 'H', y)
  y <- gsub('NN', '-', y)

  colnames(y) <- c()
  y <- data.frame(y)

  # Create the phenotype table for each sample.
  names <- y[,1]
  names <- names[-(1:3)]
  names <- data.frame(names)
  colnames(names) <- 'F2'
  names <- inner_join(names, phenotypes, by='F2')

  # Recode phenotype labels as numeric values for QTL analysis.
  # MUT and MOSAIC are coded as 0 (affected); WT is coded as 1 (unaffected).
  names$phenotype <- gsub('MUT', 0, names$phenotype)
  names$phenotype <- gsub('MOSAIC', 0, names$phenotype)
  names$phenotype <- gsub('WT', 1, names$phenotype)
  colnames(names)[1] <- 'id'
  names <- names %>% select(phenotype, id)

  # Write genotype and phenotype CSV files for the rQTL pipeline.
  outputnamegen <- here::here('data', paste0(n, '_quantgen.csv'))
  outputnamephen <- here::here('data', paste0(n, '_rqtlphen.csv'))
  write.table(x = y, outputnamegen, quote = F, row.names = F, col.names = F, sep = ',')
  write.table(x = names, outputnamephen, quote = F, row.names = F, col.names = T, sep = ',')
  return(y)
}

# Execute the conversion and also write a tab-separated QTL summary file.
x <- create_qtl_datasheet(ovule.qtl, 'ovule_f2s', phenotypes, homo1, homo2, het)
write.table(x, here::here('data','ovule.qtl.txt'), quote = F, row.names = F, sep = '\t')
```

## Build rQTL cross object and estimate map

The rQTL input CSVs are loaded into a `cross` object. Closely spaced markers are jittered to avoid zero-distance intervals, and a new genetic map is estimated from the data. The cross object and new map are saved as RDS files for use in downstream steps.

```{r rqtl,include=FALSE,eval=FALSE}
suppressMessages(library(dplyr))
suppressMessages(library(qtl))

# Create an rQTL cross object from CSV genotype/phenotype files,
# inspect the object, jitter closely spaced marker positions, estimate
# a preliminary genetic map, and save the new map for downstream steps.
ovule.cross <- read.cross('csvs', dir='',
                          genfile=here::here('data','ovule_f2s_quantgen.csv'),
                          phefile=here::here('data','ovule_f2s_rqtlphen.csv'),
                          genotypes=c("A","H","B"), alleles=c("A","B"),
                          na.strings=c("-"))
saveRDS(ovule.cross, here::here('data','ovule.cross'))
summary(ovule.cross)
plot(ovule.cross)
ovule.cross <- jittermap(ovule.cross, amount=1e-4)
newmap <- est.map(ovule.cross, error.prob=0.01, verbose=TRUE)
# Plot comparison of original vs estimated map to check for large changes.
plotMap(ovule.cross, newmap)
saveRDS(newmap, here::here('data','ovule.newmap'))
```

## Compare original and estimated maps

```{r,plot_map,echo=TRUE,eval=TRUE}
suppressMessages(library(qtl))
# Visualize the original and newly estimated maps.
oldmap <- readRDS(here::here('data','ovule.cross'))
newmap <- readRDS(here::here('data','ovule.newmap'))
plotMap(oldmap,newmap)
plotMap(newmap)
```

## Alternative map estimation via recombination fractions

As a diagnostic, pairwise recombination fractions are estimated and used to produce an alternative map. This is compared to the primary estimated map to check for inconsistencies.

```{r,est_map,echo=FALSE,eval=FALSE}
suppressMessages(library(qtl))
# Alternative map estimation using recombination fraction estimation.
# Computes pairwise RFs, plots heatmap and estimates a map from RFs.
ovule.cross <- readRDS(here::here('data','ovule.cross'))
ovule.rf <- est.rf(ovule.cross)
plotRF(ovule.rf)
newmap2 <- est.map(ovule.rf, verbose=TRUE)
plotMap(ovule.cross, newmap2)
saveRDS(newmap2, here::here('data','ovule.rf.newmap'))
```

## Replace map and calculate error LODs

The original map is replaced with the re-estimated map. Per-marker error LOD scores are calculated before and after replacement to identify potentially problematic markers. Genotype probabilities and recombination fractions are then computed on the cleaned map to prepare for QTL scanning.

```{r replace_map,echo=TRUE,eval=FALSE}
library(qtl)
# Replace the original map with the newly estimated map, calculate per-marker
# error LOD scores before and after replacement, and save summaries to disk.
newmap <- readRDS(here::here('data','ovule.newmap'))
ovule.cross <- readRDS(here::here('data','ovule.cross'))
plotMap(ovule.cross, newmap)
ovule.cross <- calc.errorlod(ovule.cross, error.prob=0.01)
errorlod1 <- data.frame(top.errorlod(ovule.cross))
write.table(errorlod1, here::here('data','ovule.toperror1.lod.txt'), quote=F, row.names=F, sep='\t')
ovule.cross.newmap <- replace.map(ovule.cross, newmap)
saveRDS(ovule.cross.newmap, here::here('data','ovule.cross.newmap'))
# Recalculate error LODs on the replaced map to identify problematic markers.
ovule.cross.newmap <- calc.errorlod(ovule.cross.newmap, error.prob=0.01)
errorlod2 <- data.frame(top.errorlod(ovule.cross.newmap))
write.table(errorlod2, here::here('data','ovule.toperror2.lod.txt'), quote=F, row.names=F, sep='\t')

# Compute genotype probabilities and recombination fractions for the cleaned map.
ovule.cross.newmap <- calc.genoprob(ovule.cross.newmap, step=1, error.prob=0.01)
ovule.cross.newmap <- est.rf(ovule.cross.newmap)
saveRDS(ovule.cross.newmap, here::here('data','ovule.cross.full.prescan'))
```

## Simulate genotypes for imputation-based scanning

Genotypes are simulated conditional on observed marker data using multiple imputation. This step is computationally intensive and is saved so it does not need to be re-run on every render.

```{r,sim_geno,include=FALSE,eval=FALSE}
library(qtl)
ovule.cross.newmap <- readRDS(here::here('data','ovule.cross.full.prescan'))
# Simulate genotypes conditional on observed data (used for imputation-based scans
# and permutation tests). This step is computationally intensive; outputs are saved
# so the simulation need not be re-run every time the report is rendered.
ovule.cross.newmap <- sim.geno(ovule.cross.newmap, step=1, n.draws=1000,
                               off.end=2, error.prob=0.001)
saveRDS(ovule.cross.newmap, here::here('data','ovule.cross.full.postscan'))
```

## Genome scan: Haley-Knott regression

A fast initial genome scan using Haley-Knott regression is run as a screen for QTL peaks before the final EM-based scan.

```{r,scanone_qtl_hk,echo=TRUE,eval=FALSE}
suppressMessages(library(qtl))
ovule.cross <- readRDS(here::here('data','ovule.cross.full.prescan'))
# Run a Haley-Knott regression scan as a fast initial screen for QTL peaks.
out.hk <- scanone(ovule.cross, method='hk', model='binary')
saveRDS(out.hk, here::here('data','out.full.hk'))
```

## Genome scan: EM algorithm with permutation test

The final method uses the EM algorithm with 1,000 permutations to determine the genome-wide LOD score significance threshold.

```{r,em_permutations,echo=TRUE,eval=FALSE}
suppressMessages(library(qtl))
ovule.cross.newmap <- readRDS(here::here('data','ovule.cross.full.postscan'))
# Run permutation test with EM scans to obtain empirical LOD thresholds.
operm.em <- scanone(ovule.cross.newmap, model='binary', method='em', n.perm=1000)
saveRDS(operm.em, here::here('data','operm.full.em'))
plot(operm.em)
# Convert permutation result object into a table of LOD thresholds and save.
lodcutoff <- data.frame(operm.em)
lodcutoff <- lodcutoff[order(-lodcutoff),]
lodcutoff <- data.frame(lodcutoff)
write.table(lodcutoff, here::here('data','lodcutoff.emscan.txt'), quote=F, row.names=F, sep='\t')
```

```{r,scanone_qtl_em,echo=TRUE,eval=FALSE}
suppressMessages(library(qtl))
# Run the final EM-based genome scan (uses imputed/simulated genotypes) and
# persist results for plotting and downstream inspection.
ovule.cross.newmap <- readRDS(here::here('data','ovule.cross.full.postscan'))
out.em <- scanone(ovule.cross.newmap, model='binary', method='em')
plot(out.em)
saveRDS(out.em, here::here('data','out.full.em'))
write.table(out.em, here::here('data','out.full.emresults.txt'), quote=F, row.names=F, sep='\t')
```

## QTL scan results

The LOD significance threshold (alpha = 0.01) is derived from the permutation distribution. A strong peak on linkage group 1 corresponds to the *BEL1* candidate region. Color and B&W versions of the plot are saved as JPEGs.

```{r save-plot, echo=TRUE,eval=TRUE}
suppressMessages(library(qtl))
out.em <- readRDS(here::here('data',"out.full.em"))
operm.em <- readRDS(here::here('data',"operm.full.em"))

lodcutoff <- as.numeric(summary(operm.em, alpha = 0.01))
cat("LOD threshold (alpha=0.01):", lodcutoff)

# Helper function to draw the QTL scan plot with LOD threshold and BEL1 annotation.
# accent_col controls the color of the threshold line and arrow (use "black" for B&W).
draw_qtl_plot <- function(accent_col) {
  plot(out.em, main='QTL Scan of mutant ovule phenotype', ylab='LOD score')
  abline(h = lodcutoff, col = accent_col, lwd = 2, lty = 2)
  arrows(x0 = 300, y0 = 10, x1 = 100, y1 = 12.5,
         col = accent_col, lwd = 2, length = 0.1, angle = 20)
  text(x = 200, y = 9, labels = "BEL1", col = accent_col, pos = 4)
}

# Display inline in the rendered report
draw_qtl_plot("red")

# Save color version
jpeg(here::here('figs',"QTLplot.jpeg"), width = 5.5, height = 4, units = "in", res = 300)
draw_qtl_plot("red")
dev.off()

# Save black-and-white version
jpeg(here::here('figs',"QTLplotB&W.jpeg"), width = 5.5, height = 4, units = "in", res = 300)
draw_qtl_plot("black")
dev.off()
```

## Two-dimensional QTL scan

A two-dimensional genome scan is run to test for pairwise QTL interactions. A coarser step size (5 cM) is used to reduce runtime. Permutation thresholds for interaction peaks are estimated with 1,000 permutations.

```{r,scantwo,echo=TRUE,eval=FALSE}
suppressMessages(library(qtl))
# Perform a two-dimensional genome scan (pairwise QTL interaction scan).
# This uses a coarser step to reduce runtime; results are saved for later use.
ovule.cross.newmap <- readRDS(here::here('data','ovule.cross.full.postscan'))
ovule.cross.err2 <- calc.genoprob(ovule.cross.newmap, step=5, error.prob=0.01)
saveRDS(ovule.cross.err2, here::here('data','ovule.cross.err2'))
out2.full.em <- scantwo(ovule.cross.err2, method="em", model='binary', pheno.col='phenotype')
saveRDS(out2.full.em, here::here('data','out2.full.em'))
```

```{r,scantwo_perm,echo=TRUE,eval=FALSE}
suppressMessages(library(qtl))
# Run permutation tests for the two-dimensional scan to obtain significance
# thresholds for interaction peaks. This is typically slow; results are saved.
ovule.cross.err2 <- readRDS(here::here('data','ovule.cross.err2'))
out2.full.em <- readRDS(here::here('data','out2.full.em'))
operm2.em <- scantwo(ovule.cross.err2, method="em", model='binary',
                     pheno.col="phenotype", n.perm=1000)
saveRDS(operm2.em, here::here('data','operm2.em'))
```

## Model comparison and QTL refinement

The two-QTL model is fit and compared against simpler models. QTL positions are refined using `refineqtl` and LOD profiles are plotted to visualize the support intervals.

```{r,compare_models,echo=TRUE,eval=TRUE}
suppressMessages(library(qtl))
# Load two-dimensional scan results and permutation thresholds, inspect
# candidate interaction peaks, fit model(s) and refine QTL positions.
ovule.cross.err2 <- readRDS(here::here('data','ovule.cross.err2'))
out2.full.em <- readRDS(here::here('data','out2.full.em'))
operm2.em <- readRDS(here::here('data','operm2.em'))
plot(out2.full.em)
summary(out2.full.em)
max(out2.full.em)
summary(operm2.em)

# Define candidate chr/pos pairs for follow-up modeling.
chr <- c(' 1',' 2')
pos <- c(35,30)
qtl <- makeqtl(ovule.cross.err2, chr, pos)
summary(qtl)
out.fitqtl <- fitqtl(ovule.cross.err2, qtl=qtl,
  formula = y ~ Q1 + Q2 + Q1*Q2,
  method  = "imp", model = "binary",
  dropone  = TRUE)
saveRDS(out.fitqtl, here::here('data','out.fitqtl'))
summary(out.fitqtl)
# Refine QTL positions and re-fit the model; plot LOD profiles.
rqtl <- refineqtl(ovule.cross.err2, qtl=qtl, formula=y~Q1+Q2+Q1*Q2, verbose=FALSE)
rqtl
plotLodProfile(rqtl)

# Fit single-QTL model for comparison
out.fq1 <- fitqtl(ovule.cross.err2, qtl=qtl,
  formula = y ~ Q1,
  method  = "imp", model = "binary")

# Compare single vs two-QTL model LOD scores
cat("Single-QTL model LOD:", summary(out.fq1)$lod, "\n")
cat("Two-QTL model LOD:   ", summary(out.fitqtl)$lod, "\n")
cat("LOD difference (interaction contribution):", 
    summary(out.fitqtl)$lod - summary(out.fq1)$lod, "\n")


```

The two-QTL model improved fit by only `r round(summary(out.fitqtl)$lod - summary(out.fq1)$lod, 2)` LOD units over the single-QTL model. Neither Q2 nor the Q1:Q2 interaction term reached significance in the drop-one analysis, consistent with a single major-effect QTL at the *BEL1* locus driving the mutant ovule phenotype.
